---
title: "Choosing ADMIXTURE k value"
author: "Katherine Drotos"
date: "2024-05-28"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```

## Introduction

This workflow is for calculating the optimal k value from ADMIXTURE runs. This method considers the original ADMIXTURE documentation, and attempts to use the Evanno et al. 2005 method of using $\Delta$k instead of the smallest value of $log Pr (X|k)$.

Example dataset: Humber leuciscid fish

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(ggthemes)
theme_set(theme_gray())

setwd("C:/Users/Katherine/OneDrive - University of Guelph/Post-doc - Mandeville/Cyprinid project/ADMIXTURE analysis/Humber_run_10_reps/")

HumberADM_CVerror <- read.table("HumberADM_allruns_cv.txt", header=F, sep="")
```

Cleaning up the dataframe to make it easier to work with:

```{r}
HumberADM_CVerror <- HumberADM_CVerror %>% 
  mutate(Run=c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,
               2,2,2,2,2,2,2,2,2,2,2,2,2,2,
               3,3,3,3,3,3,3,3,3,3,3,3,3,3,
               4,4,4,4,4,4,4,4,4,4,4,4,4,4,
               5,5,5,5,5,5,5,5,5,5,5,5,5,5,
               6,6,6,6,6,6,6,6,6,6,6,6,6,6,
               7,7,7,7,7,7,7,7,7,7,7,7,7,7,
               8,8,8,8,8,8,8,8,8,8,8,8,8,8,
               9,9,9,9,9,9,9,9,9,9,9,9,9,9,
               10,10,10,10,10,10,10,10,10,10,10,10,10,10)) # to make a Run column

HumberADM_CVerror <- subset(HumberADM_CVerror, select=-c(V1, V2)) # drop V1 and V2

HumberADM_CVerror <- lapply(HumberADM_CVerror,
                            gsub, pattern="(K=", replacement="", fixed=TRUE) # getting rid of unwanted characters

HumberADM_CVerror <- lapply(HumberADM_CVerror,
                            gsub, pattern="):", replacement="", fixed=TRUE) 

names(HumberADM_CVerror)[names(HumberADM_CVerror)=='V3'] <- 'K_value' #rename columns
names(HumberADM_CVerror)[names(HumberADM_CVerror)=='V4'] <- 'CV_error' 

HumberADMCV_df <- as.data.frame(HumberADM_CVerror)
```

Per the original ADMIXTURE documentation, the optimal value of k is determined by the lowest cross-validation (CV) error value. Here, we performed a cross validation of 10, for k=1 to k=14, repeated 10 times for each value of k (using different seeds - this must be set manually, otherwise the default seed is the same).

```{r}
HumberADMCV_df$K_value <- factor(HumberADMCV_df$K_value, 
                                 levels=c("1","2","3","4","5","6","7","8","9","10","11","12","13","14"))

HumberADMCV_df %>% 
  ggplot(aes(x=K_value, y=CV_error)) +
  geom_point()
```

The lowest CV error value here is at k=6, but there is also substantial variation across replicates above k=5. This illustrates that a single run at a single seed is insufficient, especially when there is uncertainty regarding the biological reality of the population structure (i.e., how many clusters are hypothesized to be there).

## Evanno et al. 2005 method

### Step 1: Taking the mean across runs

The first step in the Evanno et al. 2005 method is to plot the means of all runs across k values.

```{r}
HumberADMCV_df <- transform(HumberADMCV_df, CV_error=as.numeric(CV_error)) # had to make CV_error numeric

HumberADMCV_df %>%
  mutate(mean=mean(CV_error),
         .by=K_value) %>%
  mutate(sd = sd(CV_error),
          .by=K_value) %>%
  ggplot(aes(x=K_value, y=mean)) +
  geom_point() +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.2) +
  geom_line(aes(group=1))
```

Plotted another way to still see the spread of the original data (blue dots represent the means):

```{r}
HumberADMCV_df %>%
  mutate(mean=mean(CV_error),
         .by=K_value) %>%
  ggplot(aes(x=K_value, y=CV_error)) +
  geom_boxplot() +
  geom_point(aes(y=mean), color="blue", size=3)
```

### Step 2: Calculating the rate of change

From Evanno et al. 2005: "Second, we [plot] the mean difference between successive likelihood values of *K*: $L'(K) = L(K) - L(K-1)$"

In R:

```{r}
HumberADMCV_df %>%
  arrange(Run, K_value) |>
    mutate(rate = CV_error - lag(CV_error)) %>%
    mutate(sd = sd(rate),
          .by=K_value) %>%
    mutate(meanrate = mean(rate),
           .by=K_value) %>%
#    mutate(sdrate = sd(meanrate),
 #          .by=K_value) %>%
  ggplot(aes(x=K_value, y=meanrate)) +
  geom_point() +
  geom_errorbar(aes(ymin=meanrate-sd, ymax=meanrate+sd), width=0.2) +
  geom_line(aes(group=1))
```

### Step 3: Absolute values of the second order rate of change

Next, we calculate the "absolute values of the second order rate of change of the likelihood distribution" as: $|L''(K)|=|L'(K+1)-L'(K)|$

In R:

```{r}
HumberADMCV_df %>%
  arrange(Run, K_value) |>
  mutate(secondrate = abs((lead(CV_error)-CV_error) -(CV_error - lag(CV_error)))) %>% 
  mutate(srmean = mean(secondrate),
         .by=K_value) %>%
  mutate(srsd=sd(secondrate),
         .by=K_value) %>%
  ggplot(aes(x=K_value, y=srmean)) +
  geom_point() +
  geom_errorbar(aes(ymin=srmean-srsd, ymax=srmean+srsd), width=0.2) +
  geom_line(aes(group=1))
```

### Step 4: $\Delta$k

From Evanno et al., $\Delta$k is calculated as: $\Delta k = m | L''(K)| / s[L(K)]$ where $m$ is the mean of the absolute values calculated above, and $s$ is the standard deviation of the likelihood values.

```{r}
HumberADMCV_df %>%
  arrange(Run, K_value) |>
  mutate(secondrate = abs((lead(CV_error)-CV_error) -(CV_error - lag(CV_error)))) %>%  
  mutate(deltaK = mean(secondrate)/sd(CV_error),
         .by= K_value) %>%
  ggplot(aes(x=K_value, y=deltaK)) +
  geom_point() +
  geom_line(aes(group=1))
```

From this, we can see that *k*=4 clearly stands out as the selected optimal cluster model value. Coupled with the knowledge that this dataset had 4 phenotypically identified species, this makes biological sense.
